---
title: "Bayesian Data Analysis"
subtitle: "Orientation"
# For author options see : 
# https://quarto.org/docs/authoring/front-matter.html#authors-and-affiliations
# NB: multiple authors can be added here.
author:
  - name:
      given: Joshua
      family: Wilson Black
    email: joshua.black@canterbury.ac.nz
    orcid: 0000-0002-8272-5763
    affiliation: 
      - "Te Kāhui Roro Reo | New Zealand Institute of Language, Brain and Behaviour"
      - "Te Whare Wānanga o Waitaha | University of Canterbury"
format:
  revealjs:
    theme: [custom.scss]
    incremental: true
    logo: images/NZILBB-small.svg
    template-partials:
      - title-slide.html
    title-slide-attributes:
      # If you have Marsden funding, change image name to `nzilbb-uc-marsden.svg`
      data-background-image: images/nzilbb-uc.svg
      # First number controls the horizontal position, second controls vertical.
      data-background-position: '50% 5%'
      # Controls size of image relative to width of the slide.
      data-background-size: 70%
    embed-resources: false
    include-in-header:
      - text: |
          <link rel="icon" type="image/png" sizes="32x32" href=".//images/fav.png" />
bibliography: 
  - grateful-refs.bib
editor: 
  markdown: 
    wrap: 72
execute: 
  echo: true
knitr:
  opts_chunk: 
    dev: "png"
    dev.args: 
      bg: "transparent"
---

```{r}
#| echo: false
# This is a place to load packages in the background (i.e., this block won't 
# apear on the slides because `echo` is false.)
library(tidyverse)
library(here)

library(brms)

theme_set(
  theme_bw(base_size=20) +
    theme(
      panel.background = element_rect(fill = "transparent", colour = NA),
      plot.background = element_rect(fill = "transparent", colour = NA)
    )
)
```

# Overview

## Overview

- Plan for these sessions
- Fit a Bayesian model
- What's the big deal?
    - _i.e., what has changed from non-Bayesian models_
    - A _teeny_ bit about interpretations of probability.

# The plan

## Bayes sessions

::: columns

:::: {.column width="50%"}
- Four intro sessions:
  1. Orientation
  2. What does my model say? 
  3. What does my model assume? 
  4. Is my model healthy?
::::

:::: {.column width="50%"}
- Josh away: 16/03, 23/03.
- Mid-semester break
- Six sessions on specific model structuress.
    - Regression
    - Multi-level regression
    - Likert-scale data
::::

:::

::: fragment

_&c. &c. Based on your interest_.

:::


# Fit a Bayesian Model

```{r}
#| include: false
qb2 <- read_rds(here('data', 'session_1.rds'))
trap_plot <- qb2 |> 
  group_by(school) |> 
  mutate(
    student_n = length(unique(part_id))
  ) |> 
  filter(student_n >= 3) |> 
  ungroup() |> 
  group_by(part_id) |> 
  summarise(
    F1_lob2 = mean(F1_lob2, na.rm=T),
    school = first(school)
  ) |> 
  ggplot(
    aes(
      y = F1_lob2,
      x = reorder(school, F1_lob2)
    )
  ) +
  geom_boxplot() +
  theme(
    axis.text.x = element_text(size=8, angle = 45, hjust = 1)
  ) +
  labs(
    y = "TRAP F1 (Lobanov)",
    x = "High school"
  )
ggsave(
  here('plots', 'trap_schools.png'), 
  plot=trap_plot, 
  units="cm", width=16, height=9,
  bg = "transparent"
)
```

## Data {background-image="plots/trap_schools.png" background-size="contain"}

::: {.r-stack}

:::: {.fragment .backlight .centre .fade-in-then-out .r-fit-text}

"What school did you go to?"

::::


:::

## Wrangling

```{r}
#| code-line-numbers: "|2-7"
trap_sub <- qb2 |> 
  filter(
    school %in% c(
      "Avonside Girls' High School", 
      "St Margaret's College"
    )
  ) |> 
  select(
    age, school, F1_lob2, part_id
  )
```

## A model {auto-animate=true}

```{r}
#| cache: true
trap_fit <- lm(
  F1_lob2 ~ school,  
  data = trap_sub
)
```

- This is a very simple 'analysis of variance' model
- We want to see if `school` (i.e. Avonside vs. St Margarets) makes a difference to [trap]{.smallcaps} 
realisation
    - ...without thinking about confounding from other factors 😱
- How do we 'make it Bayesian'?

## A Bayesian model {auto-animate=true}

```{r}
#| cache: true
#| output: false
library(brms)
trap_fit_b <- brm(
  F1_lob2 ~ school,  
  data = trap_sub
)
```

::: {.fragment .big}

🎊 Congratulations, you're a Bayesian! 🎊

:::

- `brms` makes Bayesian methods _incredibly accessible_.
- For many models just putting in `brm()` instead of `lm()` (`lmer()`, etc...), 
will work.

##

::: {.fragment .big}
`lmer()` → `brm()`
:::

::: {.fragment .big}
'Wow! Now my model converges!' 

:::

::: {.fragment .big}

🐉🐉🐉 

:::

::: {.fragment .big}

_kia tūpato..._

:::


# What's the big deal?

## Model interpretation {.smaller auto-animate=true}

::: columns

:::: {.column width="80%"}

::::: fragment

```{r}
#| output-location: fragment
summary(trap_fit)
```

:::::

::::


:::: {.column width="20%"}

- Point estimates of coefficients
- Standard error
- Statistical significance tests

::::

:::

## Model interpretation (cont.) {.smaller auto-animate=true}

::: columns

:::: {.column width="80%"}

::::: fragment

```{r}
#| output-location: fragment
summary(trap_fit_b)
```

:::::

::::


:::: {.column width="20%"}

- 'Draws'?
- Coefficients
- Error and intervals
- 'Rhat', 'Bulk ESS'??
- No statistical significance tests

::::

:::

## Difference: no point estimates {.smaller}

- Bayesian model gives _distributions_ for all parameters.

::: {.columns}

:::: {.column width="70%"}

::::: fragment

```{r}
#| echo: false
# NB: We'll talk about tidybayes and plotting more in session 2.
library(tidybayes)
trap_fit_b |> 
  spread_draws(b_Intercept) |> 
  rename(
    Avonside = b_Intercept
  ) |> 
  ggplot(
    aes(
      x = Avonside
    )
  ) +
  geom_density(alpha=0.4, colour="darkgreen", fill="darkgreen") +
  labs(
    x = "Avonside coefficient (intercept)"
  )
```

:::::

::::

:::: {.column width="30%"}

[No privileged:]{.fragment}

- summary
- error intervals

[_`summary()` gives a mean with 95% interval._]{.fragment}

::::

:::

- Predictions from the model take into account uncertainty about coefficient.

## Difference: Bayesian CIs

- Both Bayesian and non-Bayesian models produce 'CIs', but use different 
words for the 'C':
    - '[Confidence]{.red} intervals' vs.
    - Bayesian '[credible]{.red} intervals' [(sometimes 'compatible intervals')]{.fragment}
- Interpretation: according to our data and model, there is a 95% change the
parameter is in the interval.
- There's nothing special about 95%, we can use any other number we like
(often more than one).

## 

::: fragment

```{r}
#| echo: false
# NB: We'll talk about tidybayes and plotting more in session 2.
library(tidybayes)
trap_fit_b |> 
  spread_draws(b_Intercept) |> 
  rename(
    Avonside = b_Intercept
  ) |> 
  ggplot(
    aes(
      x = Avonside
    )
  ) +
  geom_halfeyeh(
    slab_alpha = 0.6,
    fill = "darkgreen",
    .width = c(0.5, 0.7, 0.9),
    point_interval = "mean_qi",
    interval_size_range = c(2, 4)
  ) +
  labs(
    x = "Avonside coefficient (intercept)"
  )
```

:::

- This 'half-eye' plot shows 50%, 70% and 90% intervals around the mean.
- More on making these plots next time!

## Flexible estimates

- Let's predict [trap]{.smallcaps} realisation for these schools.
- Under the hood: 
    1. generate an Avonside value from the distribution, then 
    2. generate a St Margaret's value and add to the Avonside value.
- Repeat lots of times to get a distribution.
- This is easy, but the models can get much more complicated!

## [trap]{.smallcaps} realisations

```{r}
#| echo: false
predictions <- tibble(
    school = unique(trap_sub$school)
  ) |> 
  add_epred_draws(trap_fit_b) 
  
predictions |> 
  ggplot(
    aes(
      colour = school,
      fill = school,
      x = .epred
    )
  ) +
  geom_density(alpha = 0.4)
```

## _Difference_ in [trap]{.smallcaps} {.smaller}

```{r}
#| echo: false
predictions |> 
  pivot_wider(
    names_from = school,
    values_from = .epred,
    id_cols = .draw
  ) |> 
  mutate(
    Difference = `St Margaret's College` -  `Avonside Girls' High School`
  ) |> 
  ggplot(
    aes(
      x = Difference
    )
  ) +
  geom_density(alpha = 0.4, fill = "grey") +
  geom_vline(xintercept = 0, linetype="dashed")
```

- Our model and data are not compatible with saying there is no difference.
- In this case, this is pretty much identical to the distribution of the
'St Margarets' coefficient in our simple model.
- But models can get more complicated!

## Background: probability

- The alternative to Bayesianism is 'frequentism'.
- Core difference: how they use [probability]{.red}
- Frequentists: probability quantifies the reliability of the method.
- Bayesians: probability quantifies uncertainty.
- _Both are well motivated ideas!_

## Background: probability (cont.) {.smaller}

- Frequentist probability is connected to proportions of samples, e.g.:
    - 'If this analysis were repeated over and over again, we would draw the wrong
    conclusion 5% of the time.'
    - 'If you sampled NZE speakers born in 1937, 10% of would regularly use
    "kia ora" as a greeting.'
- Bayesian's use probability for uncertainty:
    - 'Given the data and model, we are 90% sure that the effect of this 
    variable is greater than zero' 
    - 'The probability that language X and language Y are in the same family is
    65%' [_(...what would you sample?)_]{.fragment}
- It's very hard to understand the Bayesian examples in terms of sampling. 
    - _'If universes were as plenty as blackberries...'_ (CS Peirce)

## Confidence vs. credible (again) {.smaller}

- Standard 95% confidence intervals: [i.e. 'frequentist']{.fragment}
    - If we repeat _this method_, the parameter we're estimating will be inside
    the confidence interval 95% of the time.
    - We're 'sampling' applications of the _method_.
    - It's not a claim about _this exact confidence interval_!
- Bayesian 95% credible intervals:
    - **Given the data and model,** we reckon, with 95% confidence, that the parameter
    is somewhere in here!
    - It's a claim about _this exact interval_.
  - Easier to interpret, but we've changed the topic.

## NB

- Bayesianism is not 'Statistics 2.0'
- There's nothing especially subjective about Bayesianism.
    - Bayesian methods _use probability_ to express some subjective elements.
    - Frequentists often put subjective elements into model assumptions etc.
- You don't have to be Bayesian
    - ...but you should understand it enough to read and review articles.

## Now and next time:

1. Make sure you have `brms` and `tidybayes` packages installed.
2. Clone or download the github repository at <https://github.com/nzilbb/ws-bayes-1>.
3. Next time: 
    - interpreting Bayesian models in more detail, esp:
    - plotting coefficients and predictions.

# References

```{r}
#| echo: false
grateful::nocite_references(
  grateful::cite_packages(output = "citekeys", out.dir = here())
)
```


::: refs

:::
